---
title: "project"
author: "Yuran Liao"
date: "2025-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev="httpgd", fig.show="hold")
```

```{r}
library(dplyr)
```
```{r}
gold <- read.csv("Gold Price.csv")
```

```{r}
gold$Date <- as.Date(gold$Date)
gold <- gold[order(gold$Date),]
```

```{r}
tail(gold)
```

Data preparation

```{r}
# check if there is missing value
sum(is.na(gold))

# detect outliers using IQR
boxplot(gold$Price, main="Gold Price Boxplot", col="lightblue")

Q1 <- quantile(gold$Price, 0.25)
Q3 <- quantile(gold$Price, 0.75)
IQR_value <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

outliers <- gold$Price[gold$Price < lower_bound | gold$Price > upper_bound]
print(outliers)

```

```{r}
plot(gold$Date, gold$Price)
```
```{r}
#gold price statistics
summary(gold$Price)
```
```{r}
hist(gold$Price)
```


From the plot of the gold price, the plot seems not stationary because the trend of the price is increasing.

Check stationary using Augmented Dickey-Fuller (ADF test):
```{r}
library(tseries)
adf.test(gold$Price)
```
By the Augmented Dickey-Fuller Test, the p-value is greater then alpha=0.05, which means we don't have enough evidence to reject that the data is not stationary. 

Make it stationary using Differencing to remove the main trend
```{r}
gold$diff_price <- c(NA, diff(gold$Price))
plot(gold$Date, gold$diff_price, type="l")
```

```{r}
adf.test(gold$diff_price[-1])
```
After differencing, the ADF test shows that the p-value is significantly smaller than the alpha=0.05, which means we can reject the null hypothesis, the data is stationary now. 








